### 2014-05-13 writeups/context-dependence.tex
edited up until summary stats

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{todonotes}

@@

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}

@@

The common framework of the above examples is a one-dimensional grid of $n$ sites,
and each site is labeled with one of a finite set of states.
These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
where each $X_i(t) \in \calS$, the set of possible states.
(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
We define a \textit{pattern} to mean a contiguous sequence of states, and we let $|u|$ denote the length of the pattern $u$.

The process starts with an arbitrary specified sequence of states $X(0)$.
The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
Thus given two patterns $u$ and $v$ of common length $h$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+h}(t)) = (u_1, \ldots, u_h)$ then
\[
\P\{(X_i(t+dt), \ldots, X_{i+h}(t+dt)) = (v_1, \ldots, v_h) = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$
determines the time evolution of the process.

It will often be convenient in the notation to avoid edge effects
by treating the $n$ sites as a circle, and treating all indices modulo $n$.
\todo[inline]{So shall we commit to this? I think that would be best, and then explain any deviation from it individually.}
\peter{I think it would be better to remove this, but I need to parse the remainder...}

@@

  This only has one parameter, the speed.
  Note, however, that it may not be entirely trivial to estimate the speed,
  since (on the circle) it may not be obvious which particle moved where.

@@

\begin{example}[CpG mutation]
  A sequence of genome can be written using A, C, G, and T;
  in the most general model of single-site mutation, each of the 12 possible transitions occurs at its own rate.
  Furthermore, it is a well-known observation that in many species, and certain contexts,
  a C followed by a G has an additional, higher rate with either the C changing to a T or the G changing to an A.
  This simplest model of context-dependent substitution rates in a genome sequence has rates

  \begin{center}
    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
      $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
      CG   &  TG   &  $\gamma$ & \\
      CG   &  CA   &  $\gamma$ &
    \end{tabular}
  \end{center}
  in which $\gamma$ is the additional CpG rate above the base mutation rate.

  This has 13 parameters.
  Note that there are no conserved quantities here, unlike in TASEP,
  which we could imagine as indestructible particles moving.

@@

@@

\paragraph{The generator matrix}
For clarity, we should define how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x$ jumps to state $y$.
Let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ be the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h, and \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+h-1$, where $x$ matches with $u$ and $y$ matches with $v$.
(Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
We'll also need the set of defined transitions in $\calT$
that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
that could change $x$ into $y$.
In principle, this gives us the transition probabilities for the process:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
but in practice the size of these matrices can make a direct application of this formula inefficient.
This paper proposes an alternate strategy.

@@

First, it will be useful to have a brief description of how to simulate such a process.
Briefly, we can put down a Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$,
and then resolve each possible change in temporal order.
\todo{OK, rad.}
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur at any position:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
\todo[inline]{OK, I'm getting confused now. From the above, I would have thought that $\mu_*$ was the overall rate of mutation, but now it's only happening from pattern x?}
\peter{$\mu_*$ is the maximum transition rate out of any possible pattern $x$.  I think the notation is correct? I have tried to clarify.}
Note that since mutation is homogeneous, the maximum rate of changes occuring at any position will be the same, except possibly near the ends of the sequence.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a rate $\mu_*$ Poisson process, and these are independent across sites.
\todo[inline]{Further confusion: now $\mu_*$ is just the rate at a single site?}
\peter{yes.}
Then, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $s_{ij} = \min_{k,h} \{ s_{kh} : s_{kh}>s \}$.
The process then jumps to a new state with probability proportional to the corresponding rate.
In other words, let
\[
q_k = \begin{cases}
  \mu_k/\mu_* \qquad & \text{if } x_i^{(|u_k|)} = u_k  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
and let $q_0 = 1-\sum_{k=1}^{n_T} q_k$.
(By construction of $\mu_*$, $q_0\ge 0$.)
Now choose transition $k$ with probability $q_k$, and if $k>0$, replace $x_i^{|u_k|}$ with $v_k$.

For instance, to simulate TASEP, one has only to let $\mu_*=\lambda$, and at each $s_{ij}$
check if at site $i$ there is a 1 followed by a 0,
and if so, switch them.

@@

The problem at hand is to infer the parameters of the model, given the state at time 0 and later at time $t$.
If $n$ is large enough and $t$ is not too large,
this should be feasible.
If $t$ is too large, the process may ``saturate'' --
if there are many changes at most sites, then in at least most models,
we will lose most of the information about the dynamics,
retaining information only about the parameters that affect the stationary distribution.
In TASEP, we lose all information as it approaches stationarity,
in the Ising model we have information about $\beta$ and $\gamma$ but not $\lambda$ from the stationary distribution,
and in the CpG model it is not immediately clear.
\todo[inline]{I would think that this is identifiable from the stationary distribution if we know the structure of the model.}

But, how can we extract the information?
These are Markov processes, on the state space $\calS^n$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible for even moderately sized $n$.
We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent, taking $m$ as large as is feasible.
Then we would define $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
denote by $x_i^{(m)}$ the block of length $m$ beginning at index $i$,
and obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km}^{(m)},y_{km}^{(m)}) .
\]
\todo[inline]{Do you prefer 1-indexing or 0-indexing for the sequences? Seems like 1-indexing before}
Variations are possible, such as modifying edge effects behavior for $p_m$.

However, in any case there is no guarantee that this is close to correct,
as it ignores dependencies between neighboring blocks.
Intuitively, we want to increase the size of the context
-- but increasing $m$ does not do this, as there are always edge effects.
\todo[inline]{I'm a little confused by this use of "edge effects." Are you using it to mean non-independence between blocks generally?}
However, we could look at the context in the initial sequence but not in the final sequence,
including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
We will call these left and right \emph{fringes}.
Concretely, define the marginal probabilities
\begin{align}
  p_{\ell,m,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_{\ell+m+r}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{\ell+m+r}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $\ell+m+r$,
then after time $t$ we will see at the middle $m$ positions the pattern $y$,
ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
\todo[inline]{Is the following notation right? It seems like taking it out is a viable option: "In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z \mid X_i^{(m)}(0) = x \}$,"}
We expect $p_{\ell,m,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large \todo{and $t$ is not} :
\begin{align} \label{eqn:window_approx}
  p_{\ell,m,r}(x,y) \approx \P\{ X_i^{(m)}(t) = y \mid X_i^{(\ell+m+r)}(0) = x \}.
\end{align}
We provide a precise statement of this approximation later.

@@

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{bm}
\usepackage{todonotes}

@@

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}

@@

Each example above describes a 1-dimensional grid of sites, each of which can take one of a finite set of states,
and that switch randomly between states according to a local set of rules.
Concretely, we suppose that we observe a finite collection of these sites at only a few times,
and that the dynamics are Markov:
with the set of possible states by $\calS$, 
and $X_i(t)$ the state of site $i$ at time $t$,
we assume that $\{X(t)\}_{t \ge 0}$ is a Markov process on $\calS^n$ 
for which the probability a given site changes state in a small amount of time
depends only on the sequence of states at nearby sites.

To make this notion of local dependence formal,
first define a \textit{pattern} to mean a contiguous sequence of states; 
let $|u|$ denote the length of the pattern $u$;
and let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the subsequence of $x$ of length $h$ beginning at location $i$.
The dynamics are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$, 
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that 
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$.
The set of all transition triples, $\calT$, determines the time evolution of the process.

A final note: recall we are focusing on finite sequences.
By default, we do not allow patterns to ``hang off'' the end of the sequence.
This is natural in many situations, but in others we might want to other boundary conditions,
modeling extant, but unobserved, neighbors.
We will mention this when necessary.

@@

  This only has one parameter, the speed.
  If we allow particles to exit from the right, and new particles to enter from the left
  (or, arrange them on a circle)
  then from observing only starting and ending configurations 
  we cannot tell which particles have moved where,
  so it is not obvious how to estimate the speed.

@@

\begin{example}[CpG mutation]
  A sequence of genome can be written using A, C, G, and T;
  in the most general model of single-site mutation, each of the 12 possible transitions occurs at its own rate.
  Furthermore, it is a well-known observation that in many species, 
  adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
  than either single nucleotide change.
  Embellishing the single-nucleotide model with this additional rate results in the model defined by

  \begin{center}
    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
      $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
      CG   &  TG   &  $\gamma$ & \\
      CG   &  CA   &  $\gamma$ &
    \end{tabular} .
  \end{center}
  Here $\gamma$ is the additional CpG rate above the base mutation rate,
  and a total of 13 parameters.
  Note that this model is very different in spirit to TASEP,
  in which the total number of particles was conserved.

  In this model, for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
  for instance, a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
  or by a $\nC\nG \to \nT\nG$ mutation; 
  and hence the total instantaneous rate at which ACG changes to ATG is $\gamma + m_{\nC\nT}$.
  Although this means there are many possible parameterizations,
  note that the model is identifiable, not overparameterized.

@@

  Glauber dynamics have been widely studied... XXX

@@

\paragraph{The generator matrix}
For clarity, we describe how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x$ jumps to state $y$.
Recall that $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denotes the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h, and \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except at positions $i,i+1,\ldots,i+h-1$, 
and in this region $x$ matches with $u$ while $y$ matches with $v$.

As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
for each position $i$, these are those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, denoted  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
that would change $x$ into $y$.
In principle, this gives us the transition probabilities for the process by a matrix exponential:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
which would then provide a route to parameter estimation.
In practice, the size of these matrices makes direct application obviously impractical.
This paper proposes an alternate strategy.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'' (also known as the Gillespie algorithm \citep{vankampen}).
Briefly, this works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, per site:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since mutation is homogeneous, the maximum rate of changes occuring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
\todo[inline]{OK, I'm getting confused now. From the above, I would have thought that $\mu_*$ was the overall rate of mutation, but now it's only happening from pattern x?}
\peter{$\mu_*$ is the maximum transition rate per site.  better now?}
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a rate $\mu_*$ Poisson process on $[0,t]$, and these are independent across sites.
\todo[inline]{Further confusion: now $\mu_*$ is just the rate at a single site?}
\peter{yes.}
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$, 
a new subsequence is then chosen at position $i$
with probability proportional to the corresponding rate:
\[
q_j = \begin{cases}
  \mu^j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{|u^j|}$ is replaced with $v^j$,
and no chance is made with the remaining probability $q_0 = 1-\sum_{j=1}^{n_T} q_j$.
(By construction of $\mu_*$, $q_0\ge 0$.)

For instance, to simulate TASEP, one has only to let $\mu_*=\lambda$, and at each $c_{ij}$
check if at site $i$ there is a 1 followed by a 0,
and if so, switch them.

@@

The problem at hand is to infer the parameters of the model, given the state at time 0 and later at time $t$.
If $n$ is large enough and $t$ is not too large,
this should be feasible -- there should be enough information in our observations to do this reliably.
If $t$ is too large, the process may ``saturate'' --
if there are many changes at most sites, then in at least most models,
we will lose most information about the dynamics,
retaining information only about the parameters that affect the stationary distribution.
In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
In the CpG model it is not immediately clear what information is retained.
\todo[inline]{I would think that this is identifiable from the stationary distribution if we know the structure of the model.}
\peter{What is identifiable?  The entire model, up to a time change?  I go back and forth on this...}

But, how can we extract the information?
These are Markov processes, on the state space $\calS^n$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible for even moderately sized $n$.
We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, there is no guarantee that this is even close to correct,
as it ignores dependencies between neighboring blocks.
Intuitively, we want to increase the size of the context
-- and the likelihood thus obtained is asymptotically correct for $m$ large,
but computationally, we are stuck with small values of $m$.

Alternatively, we could take into account context in the initial sequence only,
by including the states of an additional $\ell$ sites on both the left and right.
The probability that an initial sequence $x$ of length $m+2\ell$ after time $t$ 
is found to match a smaller, middle sequence $y$ of length $m$ is found by marginalizing:
\begin{align}
  p_{\ell,m}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^\ell} p_{2\ell+m}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{2\ell+m}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $2\ell+m$,
then after time $t$ we will see at the middle $m$ positions the pattern $y$,
ignoring what is present in the leftmost and rightmost $\ell$ positions.
We expect $p_{\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
is close to $p_{\ell,m}(x,y)$, regardless of context:
\begin{align} \label{eqn:window_approx}
  p_{\ell,m}(x,y) \approx \P\{ X_i^{(m)}(t) = y \mid X_i^{(2\ell+m)}(0) = x \}.
\end{align}
We provide a precise statement of the error in this approximation later.

@@

Equipped with marginal transition probabilities from length-$m+2\ell$ patterns to length-$m$ patterns
that correctly account for local dependencies,
we can now consider inference procedures.
